---
title: 'Lab #1 Basic R syntax/plots with data solutions'
author: "Xulin Ge"
date: "Jul 02 2023"
---


```{r, results='hide'}
# Set the directory
(WD <- getwd())
if (!is.null(WD)) setwd(WD)
```

# Read `sle_b_cell.txt` text file into R
```{r}
# using the read.table() function
txt_file <- "sle_b_cell.txt"
raw_table <- read.table(file=file.path("data", txt_file), header=T, row.names=1)
```

### Look at the dimensions of the data
There should be 26 samples.
```{r}
knitr::kable(head(raw_table), caption=" SLE B cell data set")
```

### Print the sample names to screen
```{r}
colnames(raw_table)
```

# Plot the second SLE patient sample versus the first normal control samples
Select the sample
```{r}
x1 <- raw_table["control.1"][ , 1]
y1 <- raw_table["sle.2"][ , 1]
```
### Plot in an xy scatter plot
Label the x and y-axes as 'Normal' and 'SLE', respectively.
Title the plot
Add grey grid lines with the function `grid()`
```{r}
plot1 <- plot(x1, y1, xlab="Normal", ylab="SLE",
    main="SLE B cell sample vs. Normal B cell sample – all probesets")
grid()
```

# Do the same plot but pick only the first 20 probesets
Pick only the first 20 probesets
```{r}
x2 <- x1[1:20]
y2 <- y1[1:20]
```
Use the `pch=15` argument to change the shape and color the points blue with the `col` argument
```{r}
plot2 <- plot(x2, y2, xlab="Normal", ylab="SLE",
    main="SLE B cell sample vs. Normal B cell sample – first 20 probesets",
    pch=15, col="blue") + grid()
```

# Plot the following gene in a gene profile plot
### IGLJ3 (immunoglobulin lambda joining 3), which is probeset ID `211881_x_at`.
This type of plot has the sample indices across the x-axis and the intensities on the y-axis
```{r}
# sample indices across the x-axis
x3 <- 1:26
# intensities on the y-axis cast to numeric
y3 <- as.numeric(raw_table["211881_x_at", ])
```

### First plot the ranges using the `type="n"` argument
```{r}
plot(range(x3), range(y3), type="n",
    xlab="sample indices", ylab="intensities",
    main="gene profile plot, ID = 211881_x_at") 
```

### Then add the genes with the `lines()` function call. Add grid lines
```{r}
plot(x3, y3,
    xlab="Normal", ylab="SLE",
    main="gene profile plot, ID = 211881_x_at",axes=F)
lines(x3, y3)
grid()
# las=2: labels are perpendicular(=2) to axis
axis(1, at=x3, colnames(raw_table), las=2, cex.lab=0.8)
axis(2)
```

# Boxplot
Another way to visualize a gene profile across conditions is to graph a boxplot
with a single distribution box per condition.

To do this, we need to create a factor vector that indicates the disease or normal condition like so: `f <- c(rep("SLE",17),rep("Control",9))`

Then use this vector with the expression vector for IGLJ3 in the boxplot function to
create the graph

Add points with `stripchart()` function

```{r}
f <- c( rep("SLE",17), rep("Control",9) )
boxplot(y3~f, main="gene profile boxplot, ID = 211881_x_at")
stripchart(y3~f, add=TRUE, vertical=TRUE, pch=21, bg="blue")
```


