---
title: 'Lab #3 Power and sample size'
author: "Xulin Ge"
date: "Jul 08 2023"
---

We will be working with a data set from Alizadeh et al. at Stanford [1]. 

In this study, the investigators were evaluating diffuse large B-cell lymphoma (DLBCL). Using expression profiling and hierarchical clustering (a topic that we will visit later in this class), they were able to identify 2 distinct forms of DLBCL that indicate different stages of B-cell differentiation. “One type expressed genes characteristic of germinal centre B cells (‘germinal centre B-like DLBCL’); the second type expressed genes normally induced during in vitro activation of peripheral blood B cells (‘activated B-like DLBCL’).” They also found that the germinal centre B-like DLBCL patients had a better survival rate.

We will use this data set to evaluate the power and sample size in this experiment. We will also look for the necessary number of samples to appropriately power the study.

First we will calculate the power and n required using a single gene calculation for illustration of the formula, then we will conduct a more multivariate summary that gives an idea of the power or n required for a specific percentage of genes/probes in the experiment. 

Remember that when we calculate these statistics for a microarray, we are dealing with more than a single variable, so general power formulas do not apply when attempting to summarize all genes/probes on an array.


Ref:

1. Alizadeh, A. A., Eisen, M. B., Davis, R. E., Ma, C., Lossos, I. S., Rosenwald, A., Boldrick, J. C., Sabet, H., Tran, T., Yu, X., Powell, J. I., Yang, L., Marti, G. E., Moore, T., Hudson, J., Jr, Lu, L., Lewis, D. B., Tibshirani, R., Sherlock, G., Chan, W. C., … Staudt, L. M. (2000). Distinct types of diffuse large B-cell lymphoma identified by gene expression profiling. Nature, 403(6769), 503–511. https://doi.org/10.1038/35000501


```{r, warning=FALSE, message=FALSE}
# installing/loading the package:
if(!require(pwr)) {
  install.packages("pwr"); 
  require(pwr)
}
if(!require(gdata)) {
  install.packages("gdata"); 
  require(gdata)
}
if(!require(ssize)) {
  if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
  BiocManager::install("ssize"); 
  require(ssize)
}
```


```{r, results='hide'}
# Set the directory
(WD <- getwd())
if (!is.null(WD)) setwd(WD)
```

# Read files into R
### Read Eisen DLBCL data set `eisen.txt` into R
Note this data frame contains missing values because we’re working with cDNA data
```{r}
# using the read.table() function
eisen_file <- "eisen.txt"
eisen_table <- read.table(file=file.path("data", eisen_file), 
	header=T, na.strings="NA", blank.lines.skip=F, row.names=1)
```

### Look at the dimensions of the data
```{r}
dim(eisen_table)
knitr::kable(head(eisen_table), caption="Eisen DLBCL data set")
```

### Read class label file `eisenClasses.txt` into R
```{r}
eisenclasses <- "eisenClasses.txt"
eisen_classes <- read.table(file=file.path("data", eisenclasses), header=T)
knitr::kable(t(eisen_classes), caption="eisen class label")
```

### Subset (re-index or reorder the column headers) the data frame with the class labels
If you look at the original column name order with `dimnames(dat)[[2]]` both before and after you reorder them, you will see what this has done.
```{r}
eisen_frame <- as.data.frame(eisen_table)

# look at the column name order before reordering
dimnames(eisen_frame)[[2]]

# Reorder the column headers
cl <- as.character(eisen_classes[,2])
eisen_frame <- eisen_frame[,cl]

# look at the column name order after reordering
dimnames(eisen_frame)[[2]]
```

# Pick a gene, remove cells that have "NAs", and plot the values for both classes with a boxplot and a histogram
From the eisen_classes table, we know that there are two classes in the dataset. One from 1 to 19 and the other from 20 to 39.
```{r}
# Split up classes
# Class 1
gc <- cl[1:19]
gc
# Class 2
act <- cl[20:39]
act
```

Here, we will pick the gene #7000. And split up classes and look at both samples for gene #7000
```{r}
# Pick gene #7000
# split up classes and look at both samples for gene #7000
x <- as.numeric(eisen_frame[8000, gc])
y <- as.numeric(eisen_frame[8000, act])

# remove cells that have "NAs"
x <- x[!is.na(x)]
y <- y[!is.na(y)]

# Plot the values for both classes with with a boxplot
boxplot(x, y, col=c("red","blue"), main='boxplot for eisen data gene #7000',
	axes=F, ylab="Expression")
axis(2)
axis(1, at=c(1,2), c("GC Class","ACT Class"))

# Plot the values for both classes with with a histogram
par(mfrow=c(1,2))
hist(x, col="red", main='histogram for gene #7000 GC Class', xlab="Expression")
hist(y, col="blue", main='histogram for gene #7000 ACT Class', xlab="Expression")
```


### Calculate the pooled variance
```{r, warning=FALSE, message=FALSE}
# Load the packages
library(pwr)
```
```{r}
# size of each group
nx <- length(x)
ny <- length(y)
# Calculate the pooled variance as coded in the lecture notes
pool.var <- (((nx-1)*var(x)) + ((ny-1)*var(y)))/(nx+ny-2)
pool.var 
```

### Calculate the minimum sample size necessary to detect a 1.5 fold difference (at 80% power and 99% confidence)
```{r}
dif.fold <- log2(1.5)/sqrt(pool.var)
pl.ss3 <- pwr.t.test(d=dif.fold, sig.level=.01, power=0.8, type="two.sample")
pl.ss3
```

### Calculate the sample size required for the selected gene 
Using the empirically determined delta between the two groups, assuming 99% confidence and 80% power
```{r}
# empirically determined delta
dif <- abs(mean(x)-mean(y))/sqrt(pool.var)
dif
pl.ss <- pwr.t.test(d=dif, sig.level=.01, power=0.8, type="two.sample")
pl.ss
```

The empirically determined delta shows we need at least n = 49 samples for gene #7000


# Calculate the standard deviation for each gene in the matrix 
(Hint: use the `na.rm=T` argument)
```{r}
std <- apply(eisen_frame, 1, sd, na.rm=T)
```

### plot a histogram of the standard deviations
```{r}
hist(std,n=20, col="cyan", border="blue", 
	main="Histogram of Standard Deviations for 13,412 genes", 
	xlab="Standard Deviations")
dens <- density(std)	
lines(dens$x, dens$y*par("usr")[4]/max(dens$y),col="red",lwd=2)
```


# Calculate and plot a proportion of genes vs. sample size graph 
To get an idea of the number of genes that have an adequate sample size for confidence=95%, effect size=3 (log2 transform for the function), and power=80%

```{r, warning=FALSE, message=FALSE}
# load the ssize and gdata libraries
library(gdata)
library(ssize)
```

confidence=95%, effect size=3, and power=80%
```{r, warning=FALSE, message=FALSE, results=FALSE}
fold.change <- 3.0
sig.level <- 0.05
power <- 0.8
all.size <- ssize(sd=std, delta=log2(fold.change), sig.level=sig.level, power=power)
```

Plot a proportion of genes vs. sample size graph
```{r}
ssize.plot(all.size, lwd=2, col="magenta", xlim=c(1,50))
xmax <- par("usr")[2]-1;
ymin <- par("usr")[3] + 0.05
legend(x=xmax, y=ymin, 
	legend= strsplit( paste(
		"fold change=",fold.change,",", 
		"alpha=", sig.level, ",", 
		"power=",power,",", 
		"# genes=", length(std), 
		sep=''), "," )[[1]], 
	xjust=1, yjust=0, cex=0.5)
title("Proportion of genes vs. sample size graph") 
```

The left end of the curve is too small. Enlarge the beginning part.
```{r}
ssize.plot(all.size, lwd=2, col="magenta", xlim=c(1,10), ylim=c(0,0.1))
xmax <- par("usr")[2]-1;
ymin <- par("usr")[3] + 0.05
legend(x=xmax, y=ymin, 
	legend= strsplit( paste(
		"fold change=",fold.change,",", 
		"alpha=", sig.level, ",", 
		"power=",power,",", 
		"# genes=", length(std), 
		sep=''), "," )[[1]], 
	xjust=1.9, yjust=0.5, cex=0.5)
title("Proportion of genes vs. sample size graph (enlarged)") 
```

### End