---
title: 'Lab #5 Differential expression'
author: "Xulin Ge"
date: "Jul 16 2023"
---

We will be working with a small data set run on the rat RAE230A Affymetrix array

The study was to determine differences in mRNA levels between brain hippocampi of animals fed a ketogenic diet (KD) and animals fed a control diet. “KD is an anticonvulsant treatment used to manage medically intractable epilepsies”. The differences between the 2 groups of rats can provide biological insight into the genes that are regulated due to the treatment

We are going to conduct a two-sample test for each gene/probe on the array and then identify those genes/probes that are differentially expressed between the 2 rat diet groups and plot the results with a couple of different visual summaries.


```{r, warning=FALSE, message=FALSE}
# installing/loading the package:

# No package is required in this lab
```
```{r, results='hide'}
# Set the directory
(WD <- getwd())
if (!is.null(WD)) setwd(WD)
```

# Read data set into R
Load the rat ketogenic brain data set
```{r}
# using the read.table() function
txt_file <- "rat_KD.txt"
dat <- read.table(file=file.path("data", txt_file), header=T, row.names=1)
colnames(dat)
knitr::kable(head(dat), caption="rat ketogenic brain data set")
```

# Student’s t-test
First log2 the data
```{r}
dat2 <- log2(dat)
```

From the column names listed above, we know that column 1:6 is control diet class and column 7:11 is ketogenic diet class. So, we can use the `names()` function to determine where one class ends and the other begins
```{r}
# two diet classes determined by names() function
cl <- names(dat2) 
ctrl <- cl[1:6]
KO <- cl[7:11]
```

Then use the Student’s t-test function to calculate the changing genes between the control diet and ketogenic diet classes
```{r}
# function to calculate Student’s two-sample t-test on all genes at once
# function returns the p-value for the test
# NAs are removed for each test
# s1 and s2 are dimensions of the two samples
# run function on each gene in the data frame
t.test.all.genes <- function(x,s1,s2) {
	x1 <- x[s1]
	x2 <- x[s2]
	x1 <- as.numeric(x1)
	x2 <- as.numeric(x2)
	t.out <- t.test(x1,x2, alternative="two.sided",var.equal=T)
	out <- as.numeric(t.out$p.value)
	return(out)
}

# use the Student’s t-test function to calculate the changing genes 
# between the control diet and ketogenic diet classes
pv <- apply(dat2, 1, t.test.all.genes, s1=ctrl, s2=KO)
knitr::kable(head(pv), caption="p-values using a Student's t-test")
```

# Histogram of the p-values
```{r}
par(mfrow=c(1,2))
hist(pv, col="lightblue", xlab="p-values", 
	main="Student’s t-test P-value between\ncontrol and ketogenic diets", cex.main=0.9)
abline(v=.05, col=2, lwd=2)
hist(-log10(pv), col="lightblue", xlab="log10(p-values)", 
	main="Student’s t-test -log10(pv) between\ncontrol and ketogenic diets", cex.main=0.9)
abline(v= -log10(.05), col=2, lwd=2)
```

### Report how many probesets have a p<.05 and p<.01
```{r}
length(pv[pv<0.05])
length(pv[pv<0.01])
```

5160 probesets have a p value <.05 and 2414 probesets have a p value <.01.


### Bonferroni correction - a very conservative p-value thresholding method to account for multiple testing:

Divide an alpha of 0.05 by the total number of probesets and report how many probesets have a p-value less than this value
```{r}
alpha <- 0.05/length(dat2[,1])
alpha

length(pv[pv<alpha])
```

12 probesets have a p-value less than this alpha value

### Calculate the mean for each gene
```{r}
ctrl.m <- apply(dat2[,ctrl], 1, mean, na.rm=T)
KO.m <- apply(dat2[,KO], 1, mean, na.rm=T)
```

### Calculate the fold change between the groups (control vs. ketogenic diet)
```{r}
fold <- ctrl.m-KO.m
knitr::kable(head(fold), caption="fold change between the control vs. ketogenic diet groups")
```

### Report the maximum and minimum fold change value on the linear scale
```{r}
summary(fold)

# 2^log scale value
fold.linear <- 2^fold
max(fold.linear)
min(fold.linear)

2^5.78543
2^-3.6011
```

On the linear scale, the maximum fold change is 55.15521 and minimum fold change is 0.08240443

### Report the probesets that meets: 

1. a p-value less than the Bonferroni threshold used above and 
2. |fold change|>2

Note that we were on a log2 scale for the fold change and we are looking for a linear |fold| of 2
```{r}
pv.filter <- pv[pv<alpha & abs(fold.linear)>2]
length(pv.filter)
```

There are 6 probesets meet the requirements.

### Go to NetAffx or another database source
Identify gene information for the probesets that meet the Bonferroni threshold requirements. What is the general biological function that associates with these probesets?

```{r}
names(pv.filter)
```


Search these probesets on biogps:

<br>

1367553_x_at

Hbb (hemoglobin subunit beta)

http://biogps.org/#goto=genereport&id=24440

<br>

1370239_at

1370240_x_at

1388608_x_at

Hba-a2 (hemoglobin alpha, adult chain 2)

http://biogps.org/#goto=genereport&id=360504

<br>

1371102_x_at

1371245_a_at

Hbb-bs (hemoglobin, beta adult s chain)

http://biogps.org/#goto=genereport&id=689064

<br>

We can see that these 6 probesets are all for hemoglobins (biological function) in rat, with 3 probesets mapping to the same gene

# Volcano plot
### Transform the p-value (-1*log10(p-value)) 
Make sure to use a log10 transformation for the p-value and a log2 transformation for the fold change
```{r}
# fold is already in log2 scale, transform pv
p.trans <- -1*log10(pv)
```

### Create a volcano plot with the p-value and fold change vectors

Draw the horizontal lines at fold values of 2 and -2 (log2 value=1) and the vertical p-value threshold line at p=.05 (remember that it is transformed in the plot).

```{r}
# create a volcano plot with the p-value and fold change vectors
plot(range(p.trans),range(fold), type='n', xlab='-1*log10(p-value)', ylab='fold change (log2 scale)', 
	main='Volcano Plot\ncontrol and ketogenic diets')
points(p.trans,fold,col='black',pch=21,bg=1)

# Draw the horizontal lines at fold values of 2 and -2 (log2 value=1) and 
# the vertical p-value threshold line at p=.05 (remember that it is transformed in the plot)
points(p.trans[(p.trans> -log10(.05)&fold>log2(2))],fold[(p.trans> -log10(.05)&fold>log2(2))],col=1,bg=2,pch=21)
points(p.trans[(p.trans> -log10(.05)&fold< -log2(2))],fold[(p.trans> -log10(.05)&fold< -log2(2))],col=1,bg=3,pch=21)
abline(v= -log10(.05))
abline(h= -log2(2))
abline(h=log2(2))
```


### End